---
title: 'San Francisco Multifamily Real Estate Development'
author: 'Dean Boerner/ECON 631'
date: "5/12/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width=9, fig.height=6)

library(tidyverse)
library(sf)
library("RSocrata")
library(rgdal)
library(viridis)
library(plotly)
library(ggrepel)  

#making the map theme from the textbook
theme_map <- function(base_size=9, base_family="") { require(grid)
theme_bw(base_size=base_size, base_family=base_family) %+replace% theme(axis.line=element_blank(),
axis.text=element_blank(), 
axis.ticks=element_blank(), 
axis.title=element_blank(), 
panel.background=element_blank(), 
panel.border=element_blank(), 
panel.grid=element_blank(), 
panel.spacing=unit(0, "lines"), 
plot.background=element_blank(), 
legend.justification = c(0,0), 
legend.position = c(0,0)) 
}

#reading in cleaned data
mf_raw <- read.csv('mf.csv', stringsAsFactors = FALSE)
mf_raw <- as_tibble(mf_raw)  

#converting to an sf object, where longitudes are x-coordinates, latitudes y-.
mf_sf <- st_as_sf(x=mf_raw, coords = c("Longitude", "Latitude"),
    agr = "constant",
    crs = 7131,
    stringsAsFactors = FALSE,
    remove = TRUE
    )

#Viz 1
url <- 'https://data.sfgov.org/resource/xfcw-9evu.geojson'
geo_object <- read_sf(url)
st_crs(geo_object) <- 7131

mf_joined <- st_join(mf_sf, geo_object, join=st_within)

#Vizzes 2, 3, 4
url2 <- 'https://data.sfgov.org/resource/keex-zmn4.geojson'
geo_object2 <- read_sf(url2)
st_crs(geo_object2) <- 7131
geo_object2 <- st_transform(geo_object2, 7131)

mf_supervisor_joined <- st_join(mf_sf, geo_object2, join=st_within)
```
Given data on San Francisco's real estate development pipeline, I've been tasked with finding where multifamily projects are being proposed and approved around the city, in order to inform a West Coast investment and development strategy. I'm told by my client, a real estate startup founding partner, that knowing where activity is prevalent will help them narrow their search for opportunities, as it will allow them to begin to understand whether a submarket is over- or under-saturated. Likewise, insights on approval frequency will also inform an understanding of what projects are at risk of being unapproved or delayed by neighborhood-specific contingents that can block projects.  

The data itself is posted every quarter by the city of S.F. and includes details on all active project and permit files for all types of real estate applications, including non-multifamily ones (i.e. single-family, office, retail, etc.) and applications that have been relatively dormant (e.g. included every quarter but, for whatever reason, not moving through the approval process from application to construction.) I worked with quarterly files since 2015, filtering each to only multifamily (2 or more units) projects and those having filing statuses new to a given quarter to limit duplicates.

To start, I gauge market saturation by seeing how new construction is spread across [S.F.'s 41 neighborhoods](https://data.sfgov.org/Geographic-Locations-and-Boundaries/Analysis-Neighborhoods/p5b7-5n3h).  

## Net Multifamily Units Starting Construction In S.F. Since 2015  
```{r Figure 1, fig.cap="Data Source: DataSF. Dividing number of planning approvals by applications"}
#grouping for total number of filings by neighborhood
mf_sf_summed <- group_by(as_tibble(mf_joined), nhood) %>% 
  summarize('Number_of_Filings' = n())

#adding missing neighborhoods
other_neighs <- tibble(nhood = c("Golden Gate Park", "McLaren Park", "Seacliff", "Presidio", "Lincoln Park"), 
                       Number_of_Filings = c(0, 0, 0, 0, 0))
mf_sf_summed <- add_row(mf_sf_summed, other_neighs)

#joining the map data to total number of filings on neighborhood
mf_sf_summed_joined <- left_join(geo_object, mf_sf_summed)

#checking formatting of these specific neighborhood names and making a new column in the process
mf_sf_summed_joined$Neighborhood <- ifelse(mf_sf_summed_joined$nhood == 'Golden Gate Park', 'Golden Gate Park', 
                                           ifelse(mf_sf_summed_joined$nhood == 'Lincoln Park', 'Lincoln Park', 
                                                  ifelse(mf_sf_summed_joined$nhood == 'Seacliff', 'Seacliff', 
                                                         ifelse(mf_sf_summed_joined$nhood == 'Presidio', 'Presidio', 
                                                                ifelse(mf_sf_summed_joined$nhood == 'McLaren Park', 'McLaren Park', 
                                                                       mf_sf_summed_joined$nhood)))))

#adding progress columns to the original spatially joined dataframe
mf_joined$Progress <- ifelse(mf_joined$Status %in% c('BP Filed', 'BP FIled', 'BP FILED', 'PL Filed', 'PL FILED'), 'Filing', 
                             ifelse(mf_joined$Status %in% c('BP APPROVED', 'BP ISSUED', 'Construction', 'CONSTRUCTION', 
                                                            'PL Approved', 'PL APPROVED'), 'Approval', 'Neither'))
mf_joined$Planning_Filing <- ifelse(mf_joined$Status %in% c('PL Filed', 'PL FILED'), 'Planning Filing', 'Not')
mf_joined$Construction <- ifelse(mf_joined$Status %in% c('Construction', 'CONSTRUCTION'), 'Construction', 'Not')

#filtering to only new filings, grouping by neighborhood, counting new filings, summming net units by neighborhood for those new filings, 
#and converting to tibble for join with map data joined data.
mf_other_group <- mf_joined %>% 
  filter(New_Filing == 1 & Construction == 'Construction') %>% 
  group_by(nhood) %>% 
  summarize("New_Filings" = n(), "Multifamily_Units_Started" = sum(Net_Units)) %>% 
  as_tibble()

#removing superfluous geometry column
mf_other_group <- select(mf_other_group, -geometry)

#joining new filing tibble to map data
full_other_join <- left_join(mf_sf_summed_joined, mf_other_group)
#for neighborhood labels
full_other_join$centroid <- st_centroid(full_other_join$geometry)
full_other_join$X <- st_coordinates(full_other_join$centroid)[, 1]
full_other_join$Y <- st_coordinates(full_other_join$centroid)[, 2]


title_other <- "SoMa, FiDi Have Seen The Most New Multifamily Construction"
gg_map_other_pt1 <- ggplot(full_other_join) + geom_sf(aes(fill=Multifamily_Units_Started, text=paste("<b>", Neighborhood, 
                                                                                                     ": <b>", Multifamily_Units_Started, 
                                                                                                     "multifamily units started since 2015")))
gg_map_other_pt2 <- gg_map_other_pt1 + geom_text(aes(x=X, y=Y, label=Neighborhood), size=2) + scale_fill_fermenter(palette = "YIGn", direction=1) + labs(fill = "Net Units Started", title=title_other) + theme_map() + theme(legend.background = element_blank())

ggplotly(gg_map_other_pt2, tooltip="text") %>% 
style(hoverlabel=list(bgcolor = "white"), hoveron = "fill")
```
We see that the combined area of FiDi/South Beach and south of Market Street – and particularly, the SoMa neighborhood itself – is where the greatest concentration of recent new construction has occurred. Leading the way is SoMa, which has seen more than 13,000 units start construction.  

I then look to see how effective approval processes may differ around the city by breaking new filing statuses down by supervisor district, rather than neighborhood. I also choose to see how counts compare when we control for land area to more fairly compare filing and approval tendencies across districts. Below, it shouldn't come as a surprise to see that Districts 6 and 9, which hold FiDi and SoMa, respectively, have seen the most new planning filings. After adjusting for differences in size, though, we see that District 3 (and 8, to a lesser degree) has seen a relatively large share of activity. This could be because it holds part of FiDi despite being the smallest district.

## Planning Filing, Approval Progress by Supervisor District Since 2015 
```{r Figure 2, fig.cap="Data Source: DataSF. Counting new planning applications, approvals"}
#fig 2

#defining progress in supervisor-joined data
mf_supervisor_joined$Progress <- ifelse(mf_supervisor_joined$Status %in% c('BP Filed', 'BP FIled', 'BP FILED', 'PL Filed', 'PL FILED'), 'Filing', ifelse(mf_supervisor_joined$Status %in% c('BP APPROVED', 'BP ISSUED', 'Construction', 'CONSTRUCTION', 'PL Approved', 'PL APPROVED'), 'Approval', 'Neither'))
mf_supervisor_joined$Planning_Filing <- ifelse(mf_supervisor_joined$Status %in% c('PL Filed', 'PL FILED'), 'Planning Filing', 
                                               ifelse(mf_supervisor_joined$Status %in% c('PL Approved', 'PL APPROVED'), 
'Planning Approval', 'Neither'))

#getting data to adjust for supervisor district size
geo_object2_modified <- geo_object2
geo_object2_modified$some_distance_squared <- st_area(geo_object2_modified$geometry)
geo_object2_modified$rough_square_miles <- geo_object2_modified$some_distance_squared * 3685.89
geo_object2_modified <- as_tibble(select(geo_object2_modified, supdist, rough_square_miles))
geo_object2_modified <- select(geo_object2_modified, -geometry)
geo_object2_modified$rough_square_miles <- as.numeric(geo_object2_modified$rough_square_miles)

#***test
rel_supe_by_progress <- mf_supervisor_joined %>% 
  filter(New_Filing == 1 & Planning_Filing %in% c('Planning Filing', 'Planning Approval')) %>% 
  group_by(supdist, Planning_Filing) %>% 
  summarize("Number_of_Steps" = sum(New_Filing), "Net_Units" = sum(Net_Units)) %>% 
  as_tibble()

rel_supe_by_progress <- left_join(rel_supe_by_progress, geo_object2_modified)
rel_supe_by_progress$adjusted_number_of_steps <- (rel_supe_by_progress$Number_of_Steps) / (rel_supe_by_progress$rough_square_miles)
rel_supe_by_progress$adjusted_net_units <- (rel_supe_by_progress$Net_Units) / (rel_supe_by_progress$rough_square_miles)

rel_supe_by_progress$District <- c(1, 1, 10, 10, 11, 11, 2, 2, 3, 3, 4, 4, 5, 5, 6, 6, 7, 7, 8, 8, 9, 9)

title_mod = "Districts 6, 9: Most Activity"
title_2_mod = "6, 3: Most Activity Per Square Mile"
subtitle_1_mod = "Total New Filings"
subtitle_2_mod = "New Filings Per Square Mile"


#could use seperate geom_texts with seperate agged tibbles to add labels for each fill
gross_counts_update <- ggplot(subset(mf_supervisor_joined, New_Filing == 1 & Planning_Filing != 'Neither'), 
                aes(x=reorder(supervisor, New_Filing, sum), fill=Planning_Filing))
gs2_update <- gross_counts_update + geom_bar(alpha=0.8) + 
  scale_fill_manual(values=c("springgreen3", "gold1"), labels=c("Approvals", "Applications")) + labs(title=title_mod, subtitle=subtitle_1_mod, x="District", y="Planning Filings", fill="Filing Type") + 
  coord_flip() + theme(legend.position = c(0.88, 0.82), 
                       legend.background = element_blank())

#adjusting for area
adjusted_counts_mod <- ggplot(rel_supe_by_progress, aes(x=reorder(District, adjusted_number_of_steps, sum), y=adjusted_number_of_steps, fill=Planning_Filing))
adj2_mod <- adjusted_counts_mod + geom_bar(stat = "identity", alpha=0.8) + scale_fill_manual(values=c("springgreen3", "gold1")) + guides(fill=FALSE) + labs(title=title_2_mod, subtitle=subtitle_2_mod, x="District", y="Planning Filings per Square Mile") + coord_flip()

cowplot::plot_grid(gs2_update, adj2_mod, ncol=2, align = "h")
```
  
The above also shows how often approvals come relative to filings. For a better idea of approval tendencies across districts, I finish by mapping approvals-to-applications.  

## Approvals-To-Application Ratios Since 2015 By San Francisco Supervisor District
```{r Figure 3, fig.cap="Data Source: DataSF. Dividing number of planning approvals by applications"}
supe_approvals_to_filings <- tibble(supdist=c('SUPERVISORIAL DISTRICT 1', 'SUPERVISORIAL DISTRICT 2', 'SUPERVISORIAL DISTRICT 3',
                                              'SUPERVISORIAL DISTRICT 4', 'SUPERVISORIAL DISTRICT 5', 'SUPERVISORIAL DISTRICT 6',
                                              'SUPERVISORIAL DISTRICT 7', 'SUPERVISORIAL DISTRICT 8', 'SUPERVISORIAL DISTRICT 9',
                                              'SUPERVISORIAL DISTRICT 10', 'SUPERVISORIAL DISTRICT 11'),
                                    approvals_to_filings=c((1/16), (6/31), (4/35), (1/6), (11/37), (21/70), (2/5), (4/22), (11/46),
                                                          (12/23), (1/10)),
                                    Approvals=c(1, 6, 4, 1, 11, 21, 2, 4, 11, 12, 1),
                                    Filings=c(16, 31, 35, 6, 37, 70, 5, 22, 46, 23, 10),
                                    Square_Miles=c(4.4, 5.2, 1.8, 3.8, 2.5, 3.7, 7.6, 3.4, 3.1, 8.3, 3.0))
supe_approvals_to_filings_joined <- left_join(supe_approvals_to_filings, geo_object2)
supe_approvals_to_filings_joined <- st_as_sf(x=supe_approvals_to_filings_joined, agr = "constant", crs = 7131, stringsAsFactors = FALSE)

supe_approvals_to_filings_joined$centroid <- st_centroid(supe_approvals_to_filings_joined$geometry)
supe_approvals_to_filings_joined$X <- st_coordinates(supe_approvals_to_filings_joined$centroid)[, 1]
supe_approvals_to_filings_joined$Y <- st_coordinates(supe_approvals_to_filings_joined$centroid)[, 2]

sup_title = "Multifamily-Approval Friendliness Is Highest South of Market"

sup_plot1 <- ggplot(supe_approvals_to_filings_joined) + geom_sf(aes(fill=approvals_to_filings, text=paste("District", "<b>", supervisor, "(", "<b>", Square_Miles, "sq. mi.): ", "<b>", Approvals, "approvals and", "<b>", Filings, "applications since 2015")))
sup_plot2 <- sup_plot1 + geom_text(aes(x=X, y=Y, label=supervisor)) + scale_fill_fermenter(palette = "YIGn", direction=1) + 
  labs(title=sup_title, fill="Approvals/Applications") + theme_map() 
ggplotly(sup_plot2, tooltip="text") %>%
style(hoverlabel=list(bgcolor = "white"), hoveron = "fill")
```
We see that District 10, which includes the Bayview Hunters Point and Portrero Hill neighborhoods and part of Mission Bay, has seen the highest ratio of approvals to applications. Distant runner ups are Districts 6 and 9, which include the Mission and SoMa neighborhoods.  

These and earlier findings indicate the Bayview Hunters Point and Potrero Hill neighborhoods might offer the most attractive combo of higher growth trajectory and lower disapproval risk. Though recently lower growth compared to SoMa and FiDi, the District 10 neighborhoods' proximity to higher growth areas and its own higher approval rate suggest it's a safer investment, and one that could benefit from being near SoMa and FiDi.  

For these reasons, I recommend that the firm's current property-selection efforts zero in on District 10 sites but that it hold off on any action until completing further research. More analysis is needed to understand whether the area's relative under-saturation is a flaw or a feature. Furthermore, given the time element inherent to the approval process, the above ratio might be capturing an uptick in applications ocurring before this analysis' scope leading to more current approvals, as opposed to *only* approval-friendliness. However, the gross and relative number of approvals in the area warrant serious consideration.















